---
title: "Model for age-structured malaria serology data"
author: "Nick Reich"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
```

## Problem set-up

Let $y_i$ be a a single measured marker of one particular malaria antibody from a serology run on person $i$. Additionally, let $a_i$ be the age of person $i$. Assume that all observations were collected in year $T$. We have around 5,000 observations on individuals from Lao.

We are interested in using the observed data on age and the marker to infer both a possible temporal transition from high malaria transmission in Lao to low transmission. Additionally, we would like to probabilistically classify each individual into one of two classes of ``ever infected'' or ``never infected''.

Let $f(a, t|\theta)$ be a function that takes a person's age ($a$) in  year ($t$) and computes an estimated probability that an individual was ever infected. In essence, $f(\cdot)$, is some function of the force of infection over time. Prior knowledge suggests that infection rates in Lao were high until the 1980s or 1990s, when the infection rate started to drop. The infection rate in the early 2020s is nearly zero.

A possible model for the data is:
\begin{eqnarray*}
Y_i  & \sim & \pi_i \cdot G_1 + (1-\pi_i) \cdot G_0 \\
G_0 & \sim & LogNormal(\mu_0, \sigma_0) \\
G_1 & \sim & LogNormal(\mu_1, \sigma_1) \\
\pi_i & = & 1 - \prod_{k=0}^{a_i} \left ( 1 - s(T - k) \right ) 
\end{eqnarray*}

The above model assumes that $Y_i$ is drawn from a mixture distribution where $G_0$ and $G_1$ represent the distributions of antibody values among the never infected and ever infected individuals, respectively. The probability $\pi_i$ that an individual has ever been infected is computed based on an assumed annual probability of infection (see details and derivation below). **NOTE: Maybe we want to define $\pi$ as following a Beta distribution with expectation as derived here. This would allow observations a bit more flexibility to not fit this rigid model.**

### Derivation of annual probability of infection

The discrete, annual probability of infection in year $t$ is assumed to takes values from a continuous function defined as a spline:
$$ s(t) =  B \cdot \beta. $$

This formulation makes many oversimplifying assumptions. It assumes that there is a true annual probability of infection that varied smoothly over time and that this probability is shared for everyone in the population. It parameterizes a more continuous ``force of infection'' parameter as a single discrete probability of infection in a given year (ignoring possibilities of multiple infections in a year). It assumes that the probabilities of infection from one year to the next for an individual are independent, which seems unlikely, given that some individuals may have higher risks during some times of their lives. It assumes no spatial heterogeneity across locations. 

We will specify priors on the spline coefficients that suggest the function will start at a high value and decrease to a low value, representing our prior knowledge that malaria was endemic in Lao and then became nearly eradicated by the 2020s. 

Given the above, we define

\begin{eqnarray*}
\pi_i  & = & P(\mbox{person $i$ infected at least once}) \\
 & = & 1 - P(\mbox{person $i$ never infected}) \\
 & = & 1 - \prod_{k=0}^{a_i} P(\mbox{person $i$ not infected $k$ years ago}) \\
 & = & 1 - \prod_{k=0}^{a_i} \left (1- P(\mbox{person $i$ infected $k$ years ago})\right ) \\
 & = & 1 - \prod_{k=0}^{a_i} \left ( 1 - s(T - k) \right ) 
\end{eqnarray*}

### A simulated example

This section walks through an example of simulating data and fitting a Stan model to the data. Note that the data generation model is virtually identical to the model used to fit the data. 

Start by loading the necessary packages:
```{r}
library(cmdstanr)
library(posterior)
library(splines)
library(tidyverse)

```

Read in and show the stan model
```{r, cache=TRUE}
mdl <- cmdstan_model(stan_file = "../code/serology-model.stan")
```


Here we set up a simulation to create a single dataset with 500 observations.
```{r}
N <- 500       ## num of observations
age <- rnbinom(N, mu=30, size=5)
maxage <- max(age)+1
n_basis <- 3
time_vec <- seq.int(maxage)
B <- bs(time_vec, df = n_basis)

set.seed(103)

## draw parameters from prior distributions
mu0 <- rnorm(1, log(3), 0.2)
mu1 <- rnorm(1, log(7), 0.2)
sigma0 <- rlnorm(1, 0, .1)
sigma1 <- rlnorm(1, 0, .1)
beta <- rnorm(3, c(1, 0, -3), 0.5)

## compute the transformations of parameters
Svals <- boot::inv.logit(B %*% beta)
pi <- rep(NA, N)
for(j in 1:N) {
  pi[j] = 1 - prod(1-Svals[(maxage - age[j]+1):maxage])
}

## simulate the data
z <- rbinom(N, 1, pi)
y <- (1-z)*rlnorm(N, mu0, sigma0) + z*rlnorm(N, mu1, sigma1)

```

Here are some plots to confirm that the data look like we think they should look.

For example, this plot shows the smooth function representing the "probability of infection in a given year'' as a function of time, with the year in which the study was conducted at the right hand side of the plot.
```{r}
plot(Svals, ylim=c(0,1))
```



And this plot shows that the probability of infection increases with age.
```{r}
plot(age, pi, ylim=c(0,1))
```

```{r, include=FALSE, echo=FALSE}
## for showing y as a function of age
age_cat <- cut(age, breaks = seq(0, 100, 10))
boxplot(log(y)~age_cat)
```



Now we fit the model to the data and extract some summary statistics from the fit.
```{r, results='hide'}
fit <- mdl$sample(data = list(N = N, ## num observations
                              y = y,   
                              age = age,  
                              maxage = maxage,
                              B = B),
                  chains=4,
                  parallel_chains=4)

## summarizing data and pi estimates
pi_samples <- as_draws_matrix(fit$draws("pi")) |> 
  apply(MARGIN = 2, FUN=function(x) quantile(x, probs=c(0.1, 0.5, 0.9))) |> 
  t()
colnames(pi_samples) <- c("q10", "q50", "q90")
dat <- tibble(y, z, age, pi) |> 
  bind_cols(pi_samples)

## summarizing posteriors of the group means
mu0_samples <- as_draws_matrix(fit$draws("mu0")) |> 
  apply(MARGIN = 2, FUN=function(x) quantile(x, probs=c(0.1, 0.5, 0.9)))|> 
  t()
colnames(mu0_samples) <- c("q10", "q50", "q90")
mu1_samples <- as_draws_matrix(fit$draws("mu1")) |> 
  apply(MARGIN = 2, FUN=function(x) quantile(x, probs=c(0.1, 0.5, 0.9)))|> 
  t()
colnames(mu1_samples) <- c("q10", "q50", "q90")
mu_ests <- tibble(z = 0:1, mu = c(mu0, mu1)) |> 
  bind_cols(rbind(mu0_samples, mu1_samples))
```

The following plot shows the actual probabilities of ever being infected as a function of age (black points) and the model estimated probabilities with 80% posterior intervals.
```{r}
ggplot(dat, aes(x=age)) +
  geom_point(aes(y=pi)) +
  geom_point(aes(y=q50), col="red", alpha=.1) +
  geom_segment(aes(y=q10, yend=q90, xend=age), alpha=.1, col="red")
```

And this plot shows the data (on log scale) by group (0 is never infected, 1 is ever infected), red x is the true mean for that group, the blue vertical bar is the 80% posterior credible interval for the parameter with the median at the blue hashmark.
```{r}
ggplot(dat, aes(x=factor(z))) + 
  geom_boxplot(aes(y=y)) + 
  scale_y_log10() +
  geom_point(data=mu_ests, aes(y=exp(q50)), color="blue", shape=3, size=2) +
  geom_segment(data=mu_ests, aes(y=exp(q10), yend=exp(q90), xend=factor(z)), col="blue", linewidth=2) +
  geom_point(data=mu_ests, aes(y=exp(mu)), color="red", shape=4) 
```



